---
title: "Prolonging Coal’s Sunset: The Causes and Consequences of Local Protectionism for a Declining Polluting Industry"
author: 
- Jonathan Eyer and Matthew E. Kahn^[University of Southern California]
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: [nothing, nothingness]
abstract: The share of U.S electricity generated by coal has fallen from nearly 50% to 33%.  This transition offers social environmental benefits but spatially concentrated costs as coal miners and their local communities have suffered.   Local political leaders have responded to shifting demand conditions by introducing incentives for local power plants to purchase coal from local mines. We document that power plants in ar`eas with mining activity are more likely to be coal-fired and to purchase more within political boundary coal even after controlling for the distance from power plants to mines. A one-percentage point increase in the proportion of coal plants in a NERC region with an in-state coal mine results in approximately 2.3 million additional annual tons of CO2 emissions. 
header-includes:
   - \usepackage{amsmath}
output: 
  pdf_document: 
    keep_tex: yes
linestretch: 2
number_sections: no
#bibliography: coalmining.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```


```{r, include=FALSE}
library(knitr)
read_chunk("coalmineregs.R")
```

```{r start}
```

\newpage

# Introduction
To mitigate the global challenge of climate change, nations must burn less coal. In recent years, the share of U.S electricity generated by coal has fallen from nearly 50% to 33%. The U.S reduction in coal use for generating power is especially notable because it has occurred without the U.S imposing carbon pricing or a carbon tax (Cragg et. al. 2013). The substitution away from coal is mainly due to the rise of the adoption of fracking technology and some states sharply ratcheting up their renewable portfolio standards (Venkatesh et al 2012 and Burtraw et al, 2012). While competition from natural gas is unlikely to abate, changes in federal policies such as opening additional federal lands to coal mining or rolling back carbon dioxide emissions standards on new coal plants threaten to reverse or slow this trend.


While environmentalists cheer for coal’s sunset, there are interest groups with strong incentives to protect this declining industry. Reduced power plant demand for coal imposes spatially concentrated costs borne by traditional coal mining communities in states such as West Virginia, Kentucky, and Wyoming, and the low skill workers who engage in mining and providing services in mining areas. There were 261 coal mines in the United States that shipped coal to the electricity sector in 2014. These mines are located in rural areas where the population is typically white and has little education.  Workers in these regions have fewer alternative job prospects. Their local economies experience slower economic growth and lower productivity than the rest of the nation (Islam, Minier, and Ziliak 2015 and Bollinger, Ziliak, and Troske 2011).

In this paper, we document that power plants are more likely to use coal to generate power and are more likely to purchase locally mined coal if the power plant and the coal mine are located in the same state, county, or congressional district. This finding is robust to flexibly controlling for the distance between mines and power plants.  Using a county border pairs research design, we further document the robustness of this finding.

Local politicians have strong incentives to engage in protecting local coal interests.  Coal states provide large financial incentives to encourage local coal purchases.  West Virginia provides incentives that reduce the production cost for local coal mines and encourages local buyers to buy from these sellers. Maryland and Virginia offer a \$3 per ton tax credit for utilities buying in-state coal (Bowen and Deskins 2015). Oklahoma offers credits of \$5 per ton to both coal mines and power plants, effectively contributing \$10 to every ton of Oklahoman coal that is burned for electricity generation in the state.  Both Oklahoma and Illinois have enacted laws aimed at forcing power plants to purchase their coal from in-state mines rather than cheaper coal from Wyoming.  

These local policies for protecting coal mines, their workers and their communities introduce the usual deadweight loss from subsidies.  Below, we document that the average product of labor of Appalachian coal mines – which are in regions particularly reliant on coal mining -- is much lower than in the rest of the nation.  Local subsidies represent the most likely explanation for this productivity wedge.  A second social consequence of this local protectionism is to exacerbate the social costs of power generation.

A distinctive feature of the coal industry is that its consumption and production entails large Pigouvian externalities (Davis 2011, Mueller and Mendelsohn 2012). There is deadweight loss caused by prolonging the use of a dirty technology that causes social harm. By supporting local coal mines, the transition to natural gas and renewable electricity sources is slowed and this results in additional emissions of greenhouse gases and criteria air pollutants. 

Since climate change is a global externality, local and state elected leaders have weak incentives to internalize this externality. Instead, they have strong incentives to protect local coal mines as this improves the local economy through providing high-paying jobs as well as indirectly through the local spending multiplier. We study the environmental implications of this local protectionism for local air quality and overall CO2 emissions. By quantifying the social costs associated with protection of a declining industry, this piece of our empirics is the mirror opposite of other research that attempts to measure the social benefits of protecting infant industries (see Goodstein 1995). This literature argues that society ostensibly props up infant industries because they convey a social good (Melitz 2005). 

# The Spatial Economics of Coal Trading
We present a simple framework for studying bilateral trade between power plants and coal mines. Consider the case where coal is produced by a set of spatially-differentiated mines using a homogeneous production function, and converted into electricity using a homogeneous production function by power plants that seek to minimize their cost of production subject to generating a given level of electricity.  Further, suppose there are no long-term contracts such that coal is sold on a spot market. 

Given these assumptions, we expect that a “gravity model” of bilateral trade will have significant explanatory power (Lee and Swagel 1997). If coal markets are competitive, power plants will purchase coal from the closest mine, at a price that is at most equal to the production price of coal plus the transportation cost from the second closest mine. If a power plant purchases from a mine that is not its closest prospective trading partner, it is indicative of deviation from perfect competition. 

The cost of coal transportation is a substantial portion of total coal generation costs. Shipping coal by rail – the predominant transportation method – generally costs on the order of 2-8 cents per ton per mile, so each additional 100 miles of transportation increases the delivered price of coal by \$2-\$8 per ton. Given that coal prices are generally below \$50 per ton, power plants have a substantial incentive to purchase coal from close mines. These costs are a significant portion of total operating costs for a coal-fired power plant. Based on EIA reports of average operating expenses and average heat input for coal-fired power plants, the transportation costs of moving coal 100 miles would account for about 0.2 cents per kWh, around 5% of total operating expenses.  In 2012, the median power plant purchased 960,000 tons of coal while the median coal mine delivered nearly 1.2 million tons of coal. 

Figure 1 maps the 372 coal fired power plants in the United States and the 161 coal mines in the U.S that either purchased coal in 2014 or shipped coal to the electricity sector in 2014. Several clear geographic patterns emerge. Coal mining is concentrated in Appalachia (Kentucky, West Virginia, Ohio and Pennsylvania), in southern Illinois and Indiana, and in Wyoming. While the majority of mines are in the Appalachia region, the largest mines are in Wyoming, which has relatively low-quality but easily accessible coal deposits with low sulfur content. Coal power plants exist throughout the country, but are most prevalent in the Midwest, Mid Atlantic, and South. For each power plant, we calculate the Euclidean distance to the coal mines with which it trades. Figure ~\ref{cdfdistance} presents the cumulative density function of these distances where we weight the observations by the quantity of coal the power plant consumed in 2014. We find that 30% of total power plant coal is purchased from mines that are less than 80 miles away. 

# The Empirical Strategy	
We examine the effects of political boundaries on the trading patterns of pairs of power plants and coal mines. We examine the likelihood and intensity of trading as well as contract characteristics for steam coal purchases. In each case, our key hypotheses focus on the coefficient estimates for our border variables. Our identification strategy relies on the assumption that our flexible controls for distance capture any affect associated with transportation costs between power plants and mines. 

Our sunset hypothesis posits that all else equal, there will be more coal trade when the buyer and seller are in the same jurisdiction, and that the contractual lock-in will be stronger. This econometric strategy combines the standard trade gravity model with Holmes’ (1999) borders approach. 

The distinctive feature of our econometric framework is the vector of dummy variables indicating if the origin mine and potential destination power plant share a common political jurisdiction. Power plants and mines that are within the same jurisdiction are, of course, relatively close to each other and have lower transportation costs than power plants that are far away. By explicitly controlling for the distance between power plants and mines, our political jurisdiction variable compares the likelihood of buying coal from an in-state mine relative to a similarly distance out-of-state mine.

## Coal Purchases and Quantities

We estimate a series of regressions to test the hypothesis that coal trading varies across political boundaries in ways beyond what would be predicted by a gravity model. In each case, our linear index of interest is given in equation

\begin{equation} \label{eq:baseline}
Y_{ijt}=A+g(distance_{ij})+\beta_1 Border_{ijt} +\beta_2 Power Plant Characteristics_{it}+\beta_3 Mine Characteristics{jt} + \gamma_t +U_{ijt}
\end{equation}

and we vary the definition of the dependent variable Y to reflect different dimensions along which coal mines and power plants interact. The subscript i denotes the power plant, j denotes the mine, and t denotes the year. In equation ~\ref{eq:baseline}, the key explanatory variables of interest are the vector of border dummies. We include two dummies indicating whether the mine and the plant are located in the same state or the  same Congressional District. A key point to note is that we flexibly model the role of distance on trade. This $g()$ polynomial, splines, and distance bins that we report below allow us to flexibly control for proxies for transportation costs.  

In our first set of results, the dependent variable is a dummy variable, $Trade_{ijt}$, that equals one if power plant i trades with mine j in year t. We estimate the probability of $Trade_{ijt}$ in a logistic regression, assuming that $Trade_{ijt} =1$ if $Y_{ijt}>0$.   

Similarly, we estimate OLS versions of equation \ref{eq:baseline} in which the dependent variable $Y_{ijt}$ is defined as $Quantity_{ijt}$, the amount of coal in tons that is traded between each power plant and mine combination.  

Our identification of the political boundary effect relies on the assumption that, conditional on our distance control, that a mine that is in the same state as the power plant differs from a mine that is in a different state by only the political effect. As such, we require that a minep plant combination might be either in the same state or in different states for any given distance. We limit our sample to only plant mine combinations that are greater than 10 miles apart (the minimum distance between a mine and an out-of-state plant and 335 miles (the maximum distance between a mine and power plant that are in the same state).

## A Border Pairs Test

One potential concern is that our cross-political boundary results might merely be the effect of non-linear distance effects that are not captured by our distance controls. In order to address this concern, we estimate the effect of the political boundary on coal trading in a border pairs framework in the spirit of Black (1999) and Holmes (1998). Following Dube, Lester, and Reich (2010), we rely on controls for contiguous counties that are in different states to better control for unobserved characteristics of power plant locations that could bias our estimates of the cross-state effect. Our approach relies on the typical continuity assumption that characteristics on either side of a boundary are the same. Under this assumption, the effect of the border can be treated as randomly assigned. In our case, we assume that power plants in counties on either side of a state border are otherwise comparable and then estimate the causal effect of crossing the border on the probability that a mine buys coal from a power plant. 
Following Dube, Lester, and Reich (2010), we limit our sample to only counties along state borders that are adjacent to a county in another state that also has a coal-fired power plant. There are 69 counties that both have a coal-fired power plant and are adjacent to a county in a different state that also has a coal fired power plant and a total of 89 unique county boundary pairs (some counties appear in more than one pair). There are 105 power plants in these 69 counties. Figure 2 shows the set of power plants in this sample. 
We create a set of adjacent county fixed effects for each pair of cross-state adjacent counties. Each pair of adjacent-county power plants receives a unique fixed effect. For example, there are two power plants in Clark County, Nevada and one power plant in neighboring San Bernardino County, California. Each of the three power plants receives a value of one for the Clark County-San Bernardino County dummy variable and all other power plants in the country receive a zero for this fixed effect. If a county borders two neighboring-state counties with a power plant, plants in that county would receive a one for the fixed effects corresponding to each county pairs. When we consider the likelihood that some mine in, for example, Nevada sells coal to each of these plants, the county pair specific fixed effect captures the characteristics of the region and the difference in the probability of transaction between the Nevada plants and the California plant is driven by the state border.
In order to implement this strategy, we modify equation (1) to reflect the addition of these bordering county fixed effects as 
\begin{equation}
Y_{ijpt} = \delta_p + \gamma_t + g(distance_{ij}) + \beta_1*Border_{ijt} + \beta_2*Power Plant Characteristics_{it} +\beta_3*Mine Characteristics_{jt} + U_{ijpt}
\end{equation}

Note that this new regression differs from equation (1) because we now restrict the set of observations we include in the regression and we also include a set of $\Delta_p$ county border fixed effects. Initially, we do not restrict the set of coal mining counties with which a power plant can trade, so for each of our 105 power plants we observe 400 potential trading partners in each year. We again estimate a logistic regression of the probability that a mine and a power plant trade in a given year as determined by $Trade_{ijpt}=1$ if $Y_{ijpt} >0$.

Next, we further restrict the sample so that each power plant can only buy coal from mines in its own state or in its adjacent county’s state. Figure 3 shows the set of cross-state power plant pairs in Ohio, Kentucky, West Virginia, and Indiana, as well as the mines in each of these states. Each of the power plants along the Ohio-West Virginia border have a choice set of coal mine trading partners of Ohio or West Virginia mines. Similarly, each of the power plants along the Indiana-Kentucky border have a choice set of coal mine trading partners of Indiana and Kentucky mines. As in the unrestricted border pairs framework, each pair of adjacent counties receives a fixed effect to control for unobserved regional variation. By limiting our sample to only those mines that could be in-state trading partners for at least one of the power plants in a pair, we ensure that each mine has at least one in-state potential trading partner and at least one out-of-state potential trading partner to serve as a counterfactual. 
##Contract Characteristics
In our next set of empirical results, we examine characteristics of trades. In this analysis, we restrict our study to only trades, rather than all possible power plant-mine interactions because trade characteristics like contracting status cannot be observed for trades that did not occur. We also focus on a monthly time-scale (the sharpest time-step reported in our data) rather than an annual time-scale to avoid issues associated with aggregating contract characteristics over months in a year.
We estimate a model regarding the contractual status of power plant-mine interactions. Joskow (1985) notes the importance of relationship-specific capital in the coal market, with larger quantity contracts tending to be longer duration contracts. As a corollary, we look for the presence of relationship-specific political capital in coal trades, expecting that plant-mine pairs in which political capital is prevalent or more important will be more likely to lock into long-term contracts than plant-mine pairings in which political capital does not exist. Similarly, coal mines that have guaranteed a level of demand by entering into long lasting contracts will be more likely to hire or retain workers than plants that have not guaranteed trading partners. We estimate a modified version of equation (1): a logistic regression for whether or not a trade was governed by a contract. 
## Power Plant Placement and Closures
Our gravity model estimates depend on the fact that a power plant is open at a point in time and located at a given location. These two decisions are choices and may thus be affected by access to coal mines. If coal plants are placed in coal mining jurisdictions more frequently than market forces would imply, our estimates of coal mines’ effect on local coal demand will be an underestimate of the full power exerted by mining interests. 
To study the locational choice of power plants and their choice over whether to close, we estimate a multinomial logit of whether a county has a coal-fired power plant, a non-coal fired power plant, or no power plants as a function of whether or not the county has an in-state or in-county coal mine. We assume that for a county, i, the index associated with each of these three power plant placement decisions is given as specified in equation (3).
\begin{equation}
Index_i = \beta_1 In Jurisdiction Mine_i + \beta_2 Distance to Closest Mine_i + \beta_3 Population_i + U_i
\end{equation}

We control for the Euclidean distance between the county centroid and the closest coal mine. As a result, our political boundary effects are estimated holding constant the cost associated with moving coal from the closest mine to the power plant. Because power plants are often located near population centers, we include the county’s population in the power plant siting decision.
Competition from natural gas has led to substantial reductions in coal-fired electricity capacity over the past decade. We examine the universe of coal-fired generators and estimate the likelihood that a generator is retired by 2014 as a function of whether or not the generator has an in-state coal mine.  We estimate a logit regression for whether or not a generator is retired in our sample as $Retired_i=1$ if $Y_i>0$ where
\begin{equation}
Y_i = \beta_1 Power Plant_i + \beta_2 In Jurisdiction Mine_i + \beta_3 Distance to Closest Mine_i + U_i 
\end{equation}

If the coefficient on our In Jurisdiction Mine coefficient is less than zero, power plants with an in-state coal mine are less likely to retire than plants without an in-state coal mine, even after controlling for the distance to the geographically closest potential trading partner. 

# Power Plant and Mine Data
In this study, our main unit of analysis will be a coal mine’s trading with each power plant in each year.  The EIA collects data on fuel deliveries to the plants in the power sector. Since 2008, fuel deliveries have been collected on the EIA-923 form, which covers monthly fuel deliveries by both utilities and non-utility deliveries. Between 2002 and 2008, utility fuel deliveries were reported through the FERC-423, while non-utility deliveries were collected via survey with the EIA-423 for plants in excessive of 50 MW. Prior to 2002, only the FERC-423 existed, and collected only utility fuel purchases.
Since the earlier data do not report the mine-specific MSHA ID, we treat a mine’s location as the mine’s county of origin, which is reported much more consistently and aggregate coal deliveries to the plant-coal county-year (aggregating across months in the year). We then construct a data set containing 5,289,000 rows. Our unit of analysis is a coal mine/power plant/year.  In our data set, there are 410 coal mine counties, 516 power plants and 25 years (1990-2014) in our sample. For each mining county, we calculate the latitude and longitude of the county centroid and compute the distance between the county centroid and each power plant. Finally, we calculate the total annual quantity of coal that is shipped for each county and the total annual quantity of coal that is received for each power plant and drop plant-coal county-year observations in which the plant received no coal or the coal county shipped no coal. This leaves 1.9 million observations from the initial matrix of 5.2 million plant-mine-year combinations. We supplement the EIA/FERC data with additional power plant-specific characteristics such as the utility type of the power plant operator (e.g. municipally-owned, investor-owned, independent power producer, etc). 
In the recent EIA sample (2008-2014), characteristics of the delivered coal are also reported for each transaction. These characteristics include the delivered cost of the coal , as well as the ash, sulfur, and heat content for the fuels. The more recent data also report characteristics of the trade such as whether or not the trade occurred pursuant to a contract and the duration of any contracts. This sample also contains the mine-specific MSHA ID of each coal mine rather than only the county of origin. We exploit the increased precision of coal mine locations in this data and compute the distance between each mine (rather than the county centroid of the mine) and each power plant based on the reported latitudes and longitudes of each mine and plant. We then overlay state, county, and congressional boundary shape files onto our geospatial data on power plant and coal county/mine location and create indicator variables for whether a power plant – coal county/mine combination are in the same state, county, or congressional district.
#Results on Coal Trading
Table 2 reports estimates of equation (1) and shows that coal purchases are generally local. Columns 3 and 4 show the percentage of coal deliveries to power plants in each state that were from in-state and out-of-state coal mines, respectively. Columns 5 and 6 show the percentage of coal deliveries from mines in each state that were to in-state and out-of-state power plants, respectively. Columns 4 and 5 are blank if a state does not have any coal mines that shipped to the power sector throughout the duration of our sample.
The average state receives 24% of its total coal consumption from in-state mines, although this is biased downward because many states do not produce coal at all and must purchase all of their coal from out-of-state mines. The average across only the states that produce coal is 48%. The states that receive the lowest-percentage of total coal purchases from in-state mines were Kansas (20), Maryland (24), Missouri (29), Oklahoma (40) and Tennessee (47). In each of these states, coal is the dominant source of electricity generation but coal mining is a relatively small industry.

##Coal Purchases and Quantities Results
The prevalence for intra-state trades is not driven merely by distance. Across each distance specification, we consistently find a positive and statistically significant effect of being within state lines on the probability that a power plant will purchase coal from a mine. Table 3 presents these results. A power plant is 0.3 percentage points more likely to purchase coal from an in-state mine than from an out-of-state mine that is the same distance away. This effect is quite large in context. Across our entire sample, the probability that a plant-mine combination engages in a trade in a given year is about 2 percent. We find even stronger effects at the county boundary. Intra-county trades are approximately 2 percentage points more likely to occur than inter-county trades. The effect of the congressional district is comparable in magnitude to the effect of the state boundary. These effects are consistent regardless of the approach to controlling for transportation distance using either polynomials or a restricted cubic spline. Using the 10-mile bins – the finest granularity of distance control – the state effect is reduced slightly. Mines and power plants that share a state are 0.2 percentage points more likely to trade than those that are in different states. The effect of the county boundary is reduced by an order of magnitude, and the statistical significance is weaker for both the county and congressional boundary estimates. Also, note the relationship between the political boundaries. A mine that is in the same county and congressional district as a power plant is obviously in the same state as well, so the net effect is the sum of the three coefficients.
The probability that a power plant and a mine trade is increasing in both power plant purchases and in mine shipments. This indicates that plants that buy a lot of coal tend to purchase from more mines than plants that buy a relatively small amount of coal. Similarly, mines that produce a lot of coal sell to more power plants than small mines. 
We find similar results when we examine the joint distribution of a trade occurring and the amount of coal that is purchased. These results are reported in Table 4. A power plant will buy approximately 8,000 more tons of coal from an in-state mine than it would buy from an out-of-state mine of comparable distance. The effect of the congressional border is approximately 100,000 tons, around eight times the magnitude of the cross-state effect. The effect of the county border is substantial; a mine will buy 1.4 million more tons of coal from an in-county power plant than it would buy from an out-of-county power plant. In each case except the cross-state effect in the splined distance control and the state and congressional boundary effect in the ten-mile bins, the estimates are strikingly similar across distance controls.

##Border Pairs Results
Our border pair models presented in Tables 5 and 6 report further evidence of political boundary effects in coal purchasing. In estimating equation (2), we include all mines as potential trading partners, each of the 105 power plants in the 69 counties that both have a coal-fired power plant and border a county with a coal-fired power plant can trade with any of the 400 mines.
We find that plants are approximately 4 percentage points more likely to purchase coal from in-state mines that from mines in a different state. Again, this effect is robust to a range of controls for the distance between the power plant and the coal mine. The same state boundary also corresponds in some cases to a shared congressional district and county which we omit from the estimation.  Using the border pairs subsample, we find larger overall effects than our baseline specification. The combined effect of the three boundaries in our baseline results is around 2-3 percentage points.
Next, we further restrict our sample by limiting each pair of adjacent-county power plants to only being able to purchase coal from mines in either state of the adjacent counties. For example, the power plants in Mobile County, Alabama and Jackson County, Mississippi would have as potential trading partners the coal mines in Alabama and Mississippi but we drop the power plant-mine observations in which the mines are in other states. In this specification, the border variable captures the relative difference in the probability of a trade between each mine in Alabama and the plants in Mobile, County Alabama and Jackson County, Mississippi. Our neighboring county-pair fixed effect captures unobserved characteristics of the Mobile/Jackson area and the remaining difference in probability is assigned to the border effect. The results are slightly larger in the specification that limits a power plant’s potential trading partners to only the in-state mines on either end of the border. In this specification, the effect of a sharing a state is in the neighborhood of 5 percentage points. 
##Contract Characteristics Results
When we focus on the contracting characteristics of the trades we substantially reduce our sample size because we do not observe trade characteristics for trades that do not occur. After focusing only on trades that occur, the correlation between our cross-boundary variables increases because many of the, for example, different-county but same-state observations that induced orthogonality between the political boundary controls did not result in trades. As such, we highlight the results of F-tests for the joint significance of our political-boundary variables in each of these cases. 
In Table 7 we estimate the effect of our boundary effects on the probability that a trade is associated with a long-term contract rather than a spot trade. Each of our political boundary controls is positive but statistically insignificant when we examine whether power plants and mines enter into contracts or interact on the spot market. When we limit our regression to only the state-level boundary effect, we find a statistically significant boundary effect indicating that intra-state trades are less likely to be spot trades than cross-state trades. 
These results are broadly consistent with state specific/relationship-specific capital being present. Joskow (1984) who attributed the longer duration of relatively high quality contracts to relationship-specific capital. Moreover, the effect of the boundary control on contracting prevalence indicates the potential for a different type of relationship specific capital. While Joskow (1984) focused on the physical characteristics of the plants and the coal that they burned, our result is consistent with relationship-specific political capital inducing power plants and mines to operate together.

## Power Plant Placement and Closures Results
In Table 8, we report estimates of equation (3).  We find a political jurisdictional effect in power plant placement decisions. After controlling for the distance to the closest coal mine, the probability of having a coal fired power plant in a county is about 0.5 percentage points higher if there is a coal mine in the county than if there is not a coal mine in the county. Note that this indicates that even after controlling for the distance to the closest coal mine, a county is more likely to have a coal burning power plant if there is an in-county coal mine. Surprisingly, the effect of having an in-state but out-of-county coal mine actually serves to decrease the probability that there is a coal power plant in the county. This effect is quite small relative to the in-county effect.
In Table 9, we find a jurisdictional effect in coal power plant closures. A power plant (or rather a generator) that is in a state with a coal mine is approximately 7 percentage points less likely to have closed by 2014 than a coal power plant without a potential in-state trading partner. In each case, these results suggest that coal mines’ political influence on the electricity sector exceeds our primary estimates which take coal plant stock as exogenous.

#The Social Cost of Within Coal Mining States Transactions
Coal burning features a high carbon intensity of roughly 2300 pounds per MWH of power. In addition to producing this global externality, coal use by power plants creates many local negative environmental and health externalities. These externalities are shown to reduce nearby home values (Davis 2011). In an area suffering from high unemployment, these may be secondary concerns.


We estimate aggregate emissions from the electricity sector due to coal protection. In order to examine total emissions, we obtain CO2, SO2, and NOx emissions from the EPA’s Air Markets Program Data (AMPD). AMPD reports emissions for each power plant that is covered under any air markets program. Most power plants are covered by at least one air markets program in the latter half of our sample. We then aggregate CO2, SO2, and NOx emissions across power plants in each NERC region up to the monthly level.
We estimate

\begin{equation}
Y_{nt} = \beta_1 In Jurisdiction Mine Percentage_{nt} + \beta_2 NERC Characteristics_n + T_t + U_{nt}
\end{equation}

where Ynt is the CO2, NOx, and SO2 emissions from AMPD regulated power plants in nerc n in year t, In Jursidiction Mine Percentage is the proportion of coal-fired power plants in a NERC region that have an in-state coal mine and NERC Characteristics is a matrix of NERC characteristics. 
If coal power plants with nearby coal mines are more likely to continue to operate – or operate at higher levels of generation – than coal power plants without nearby coal mines, we would expect a positive coefficient on the In Jurisdiction Mine Percentage variable. 
Table 10 presents these regression results.  We find that the percentage of power plants in a NERC region that have an in-state coal mine trading option leads to increases in CO2, SO2, and NOx emissions. In the case of CO2, for every percentage point increase in the proportion of power plants in a NERC region that are close to an in-state mine, an extra 2.3 million tons of CO2 is emitted. Assuming a social cost of carbon of \$40 per short ton, these results suggest that each percentage point increase in the number of coal-fired power plants in a NERC region with a potential in-state trading partner results in \$92 million per year in added social costs. Based on the average prevalence of in-state mines in our data set (53%), this would indicate that around 8% of CO2 emissions from the electricity sector are attributable to this effect.

#Local Effects of Within-State Coal Transactions
Mining communities that can rely on politically-motivated power plants to buy their coal should differ from mining communities that do not have the opportunity to “capture” power plants. Such mining communities should have lower unemployment and higher wages than a similar community with fewer intra-jurisdictional potential trading partners. Moreover, because income and unemployment are highly correlated with life expectancy (see, e.g. Chetty et al, 2016), people in areas with the potential to capture a power plant should live longer than those in areas with fewer intra-jurisdictional plants.
                                                                                 Similarly, mines that can rely on political power to find buyers for their coal should be less productive than mines that face competitive market forces. In a national competitive market for coal, the marginal product should be equal across all mines. When some power plants deviate from the competitive model and prop up local trading partners, however, these mines will over produce relative to the competitive case, hiring additional and marginally less productive workers.
                                                                                 When estimating the local effects of coal protectionism, we focus on the effect on counties and mines of having access to in-state trading partners. Because coal-mining regions, in general, are poorer than non-coal mining regions it is important to control for the underlying exposure to coal power plants. This separates the general effect that coal consumption tends to be high in poor regions from the specific effect that intra-jurisdiction coal consumption can provide an excess benefit.
                                                                                 Our regression equations are of the form 
                                                                                 \begin{equation}
                                                                                 Y_{it}= \beta_1 Close Coal_{it}  + \beta_2 Close In State Coal_{it}+Entity Characteristics_{it}+ \epsilon_{it}
                                                                                 \end{equation}
                                                                                 Our dependent variable, $Y_{it}$, is a set of county characteristics including mortality rates, unemployment rates, and average annual salaries.  The key control variable and the variable of interest, Close Coalit and Close in State Coalit, respectively, are defined as the quantity of coal consumed in the electricity sector within 100 miles of a county. The former variable reflects the generally weak economies of coal country. The latter variable captures the differential effect of being in coal country, but having access to intra-jurisdictional coal purchasers. In the county-level outcome regressions, Entity Characteristicsit, is a matrix of county information that affects local economic and social outcomes. In the mine-level productivity regressions it is a matrix of mine information that affects productivity.
                                                                              
In each of the county-specific outcomes regressions presented in Table 11, we find that counties in mining regions with access to in-state coal-fired power plants have  better outcomes to counties with less access to in-state plants. Each additional million tons of coal that is consumed by in-state power plants within 100 miles results in 1.1 fewer deaths per 100,000 residents, a 0.04 percentage point reduction in the unemployment rate, and \$76 in additional average earnings. In each case, the effect of coal consumption by all nearby plants (in-state and out-of-state) is the opposite sign, indicating that the in-state purchases are offsetting the disamenity associated with being in a coal-heavy area. 
                                                                                 #Mechanisms
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         Our main explanation for within political boundary trading focuses on political intervention. Elected officials such as a mining state’s governors, Congressmen and local officials have an incentive to help their constituents. Miners and the members of their communities are typically low skill people with long time roots to the area. Local elected officials in coal states are aware that their constituents face significant dislocation costs and seek to protect them from long-lasting negative income shocks by stabilizing demand for their constituents’ output. The conversion of a power plant in Kentucky from coal to natural gas, for example, resulted in substantial opposition from state officials representing the area on the grounds that the reduction in coal demand would harm the local economy. 
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         Elected officials in coal areas have strong incentives to take actions that increase the demand for coal. Mining is a high paying job for low skill workers. Indeed, the average wage for all U.S. coal miners was \$82,000 in 2013 according to the National Mining Association. The average wage across all industries, by contrast, was under \$50,000. Coal mining counties tend to have a slightly weaker economy than non-coal mining economies. In 2014, the average unemployment rate in counties with coal mines was 6.8 while the average unemployment rate in counties without coal mines was 6.5.  Because coal mining activity is concentrated in relatively rural communities and require relatively little education, these statistics actually underestimate the coal mining wage premium. Further, mining jobs create a local multiplier effect in their community, resulting in additional job opportunities and economic activity.
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         By boosting coal demand and raising local wages and home prices, elected officials can achieve stability for local families. Stable families and their communities go hand in hand in areas that do not have alternative industries to turn to. Families enjoying job security are less likely to experience divorce, substance abuse and economic hardship. Labor economists have emphasized the possibility of scaring due to duration dependence associated with experiencing unemployment (Borjas and Heckman 1980, Black and Sanders 2002, Black, McKinnish and Sanders 2003, 2005a, 2005b). If coal miners lose their jobs and become unemployed, the duration dependence hypothesis posits that they will become increasingly less likely to find a new job and are unlikely to find a new job that pays as well.  Urban economists have emphasized the importance of durable real estate capital in determining local real estate prices (Glaeser and Gyourko 2005).  Given the inelastic supply of housing in Appalachia, reductions in local demand for housing near declining coal mines would result in falling home prices. This would mean that unemployed coal miners who own homes would suffer the double blow of not being employed and owning an asset that has declined sharply in value.  
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         The children who grow up in such families are likely to be especially affected. Based on Heckman’s dynamic complementarity model, the early years of life are crucial for raising the chances of a child achieving her full potential (Heckman 2007). If the household’s income declines and if the family divorces, such a child is less likely to succeed. This sad dynamic shows how reduced coal demand translates into widening income inequality and increased poverty in these rural areas.
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         Elected officials in coal areas will understand this dynamic and this creates an incentive (due to both altruism for constituents as well as the desire to be re-elected) for local politicians to use their clout to encourage in state coal power plants to “buy locally”. In this sense, local elected officials internalize the social benefits from boosting coal demand but have only weak incentives to internalize the social costs of coal burning.  
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         There are alternative mechanisms that could contribute to within jurisdiction coal trade. Past investments by local power plants to optimize the power generation process as a function of local coal purchases may create a “lock in” effect through the asset specificity of past investments (Joskow 1985). One possible explanation for why the two parties would be willing to “lock in” to a long term relationship is because they anticipate that there will be less future political risk between two contracting parties since they are both represented by the same political leaders. 
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         To provide some evidence on the relevance of different potential mechanisms for understanding “within border” coal trade, we study how the propensity to purchase in state coal grows with state leverage over power plants and in times of recession. One might expect that elected officials would be more likely to use their clout to encourage the demand or in state coal mining when the economy is weak.
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         Government-owned utilities would be more susceptible to job protectionary pressure than independent power producers who would be more concerned with profit maximization and less concerned with local stakeholders. We would therefore expect a smaller border effect when a greater proportion of power plants were controlled by independent power producers. Cicala (2014) documents that when utilities were forced to divest their power plants to new owners during electricity deregulation that fuel procurement costs declined for coal plants. If political protectionism is responsible for some of Cicala’s (2014) effect, we would expect our border protection estimate to be smaller after deregulation than before it.
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         In order to investigate this dynamic over time we estimate our primary specification as a cross section for each year between 1990 and 2014, dropping the time specific fixed effects. Estimating the model separately for each year, rather than estimating a single model with interactions between the time fixed-effects and the political boundary variables lets our distance control adjust over time so that changes in the political protectionism over time are not confounded with changes in shipping infrastructure or cost. Figure 4 shows the 95 percent confidence interval on the same state coefficient over time (the same Congressional District and same county patterns yield similar findings). From 1990-1995, the coefficients increased indicating relatively greater protectionary behavior. Between 1995 and approximately 2000, there was relatively less protectionary behavior. Protectionary behavior was relatively consistent between 2000 and 2006 after which it generally decreased through 2014.
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         The majority of our sample falls after electricity deregulation was already in effect or being implemented so we do not have pre-deregulation coefficients with which to compare against the early 1990s results. Still, it is surprising that protectionist behavior is growing during the expansion of deregulation during the early 1990s. This provides some suggestive evidence our political jurisdiction effect is not driving the reductions in procurement costs noted by Cicala (2014). 
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                         We also note that changes in political climate throughout our sample period. The magnitude of the boundary effect is generally decreasing (less local purchasing) during the 1990s, the late 2000s and the 2010s, while it is increasing (more local purchasing) during the early and mid-2000s. Note that the 1992-2000 and 2008-2014 periods with declining local protectionary power aligns with a Democratic Presidential Administration, while the increasing protectionary power of the 2000s aligns with the Republican Bush Administration. We do not, however, attempt to assign any causality between the political climate and the effect.
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                 In Table 12, we present the average productivity of each coal basin between 2001 and 2014 (excluding 2012) as measured in tons of coal per labor hour, weighted by the number of employees at each mine. Productivity is lower in Appalachia than in other coal mining regions. While average productivity has fallen in each mining basin, the decline in average productivity over time has fallen quickly in Appalachia relative to most other regions. One possible explanation for this is that Appalachian mines receive relatively high prices for their coal as a result of political subsidies or political pressure.

#Conclusion 
The phase out of coal would be a Hicksian Pareto improvement for the United States. Based on a Social Cost of Carbon of \$30 a ton, the current national externality from coal associated with in-state mine purchasing (abstracting from criteria air pollution costs) was \$4 billion in 2014.  While it may be socially efficient to phase out coal, the costs of such a transition are spatially concentrated. 
                                                                                 States that specialize in mining have incentives to promote the growth (or at least slow the decline) of one of their key industries. Given the durability of housing capital and the built up social networks established in mining areas, its residents face both migration costs and asset losses if the demand for coal mining declines. Such individuals face a fundamental job retraining challenge that middle-aged workers who have worked in mines will have trouble transitioning to other jobs. In this sense, there are local “Social Benefits of Carbon” and these benefits have been ignored in studying issues related to the political economy of phasing out coal. Local officials in coal areas are well aware that many of their constituents depend on the continuing viability of the coal industry. Local officials internalize the benefits of coal’s prolonged sunset but they ignore the social environmental costs associated with such implicit subsidies. Our examination of the sunset of the coal industry revisits recent research that has examined how rural communities have gained from the fracking boom (Feyrer, Mansur, and Sacerdote (2015) and Alcott and Kenniston (2014)). 
                                                                                 While the Trump administration has signaled a willingness to roll back environmentally-motivated restrictions on coal mining and coal burning, these changes will likely coincide with less-restricted production of natural gas. Because natural gas is economically competitive with coal for baseload electricity generation, the new administration may in fact weaken demand for coal in the U.S. Concurrently, the removal of social safety net schemes and a potential shift of the tax burden from the rich to the poor would further weaken these communities. This would suggest that the ability of local politicians to prop up local coal demand will grow even more important in the coming years.
                                                                                 We have introduced a detection approach to measure “excess” within border transactions. Our empirical research design exploits the fact that coal mines and power plants vary with respect to their geographic location. Some lie within the same political jurisdiction while others do not. This variation allows us to use a flexible distance polynomial between pairs of power plants and coal mines to recover key border effects. We document an increased likelihood of trade and larger trade quantities when partners are within the same state, county, and congressional district. 
                                                                                 The costs of the CO2 emissions associated with this local “excess demand” are substantial, nearly \$100 million a year in excess social carbon costs. Still, sustained low natural gas prices may limit the duration of coal’s sunset as market forces drive additional natural gas capacity to supplant coal plants. Given the proximity of environmental tipping points associated with particular levels of CO2 concentrations as well as uncertainty about energy policy under the Trump Administration, coal protectionism may still cause major environmental costs.
                                                                                Future research could consider how to design a coal buyout that compensates coal miners and their communities for their expected losses. Harstad (2012) has proposed a mechanism of “buying coal” from marginal unregulated coal producers and then keeping these purchased deposits in the ground after restricting production among regulated coal producers. While Harstad focused on the international problem of unregulated coal consumption, his proposal could be modified to address the problem of coal’s sunset by purchasing domestic coal reserves and by paying coal miners to mine the purchased coal. The coal could then be stored rather than burned, providing the local economic benefits associated with the coal mine while avoiding the increased electricity costs and environmental externalities caused by the combustion of the coal. By paying the coal miners and supporting the local economy, local politicians would no longer need to use their political capital to incentivize continued local coal consumption.
                                                                                
```{r reg1}
```

```{r reg2}
```